---
title: fn_compare_expression_matrices
author: Haky Im
date: '2020-12-03'
slug: im-compare-expression-matrices
categories: []
tags: []
description: ''
topics: []
editor_options: 
  chunk_output_type: console
---

Want to write a function that takes in two data frames with expression data and returns 

- vector of correlations of matched genes
- summary of the correlation vect (min,1Q,median,3Q,max)
- returns a list of aligned data frames
- plots correlation for a sample of matched genes
 
- print number of individuals in df1 and df2
- print number of individuals in common between df1 and df2
- print number of genes in common between df1 and df2
- print number of genes in df1 not in df2
- print number of genes in df2 not in df1

```{r, echo=FALSE, eval=TRUE }
suppressMessages(library(tidyverse))
##suppressMessages(install.packages("vroom"))
suppressMessages(library(vroom))
work_dir = "/Users/haekyungim/Box/LargeFiles/GTEx/tests"
##system(glue::glue("scp criviarcc:/gpfs/data/im-lab/nas40t2/natasha/predixcan/psychencdoe_hg19__predict_new.txt {work_dir}"))
##system(glue::glue("scp criviarcc:/gpfs/data/im-lab/nas40t2/natasha/predixcan/amygdala__predict_new.txt {work_dir}"))
#df1 = read_tsv(glue::glue("{work_dir}/psychencdoe_hg19__predict_new.txt"))
#df2 = read_tsv(glue::glue("{work_dir}/amygdala__predict_new.txt"))

df1 = vroom(glue::glue("{work_dir}/psychencdoe_hg19__predict_new.txt"))
df2 = vroom(glue::glue("{work_dir}/amygdala__predict_new.txt")) 

```

Comparison function

```{r, warning=FALSE}

fn_remove_ver_from_names = function(df)
{
  ind = substr(names(df),1,4) == "ENSG"
  names(df)[ind] = substr(names(df)[ind],1,15)
  df
}


fn_compare_expression_matrices = function(df1, df2, nplots = 4, common_genes=TRUE,plot_dir=NULL)
  {
  ## this function assumes that df1 and df2 have the first two columns named FID and IID"
  ## usage: 
  ## tempo_list = fn_compare_expression_matrices(df1, df2, plot_dir="working_dir/plots")
  ## to see plots on screen use the default value plot_dir=NULL by not specifying anything
  ## tempo_list = fn_compare_expression_matrices(df1, df2)
  print(dim(df1)) 
  print(dim(df2))
  if(!identical(df1$FID, df2$FID) |  !identical(df1$IID, df2$IID) )
  {
    id_tibble <- full_join(hg19[,1:2], amygdala[,1:2])
    print(paste("in total there are ",nrow(id_tibble) , "rows") )
    df1 = left_join(id_tibble, df1, by=c("FID","IID"))  
    df2 = left_join(id_tibble, df1, by=c("FID","IID"))  
    print(identical(df1$FID, df2$FID))
    print(identical(df1$IID, df2$IID)) 
    print("the two lines above should say TRUE")
  } 
  ## remove version number from ensid
  df1 = fn_remove_ver_from_names(df1)
  df2 = fn_remove_ver_from_names(df2)
  if(common_genes)
  { 
    gene_list = intersect(names(df1),names(df2))
    gene_list = gene_list[!(gene_list %in% c("FID","IID"))] ## remove FID, IID
    print(glue::glue("There are {length(gene_list)} genes in common"))
    cor_vec = rep(NA, length(gene_list)); names(cor_vec) = gene_list
    for(gg in gene_list)
    {
      cor_vec[gg] = cor(df1[[gg]],df2[[gg]])
    }
    print(summary(cor_vec))
    for(pp in 1:nplots) {
      gg = sample(gene_list,1)
      if(!is.null(plot_dir)) png(glue::glue("{plot_dir}/{gg}_df1_vs_df2.png"))
      plot(df1[[gg]],df2[[gg]],main=paste(gg," - cor = ",round(cor_vec[gg]),2))
      if(!is.null(plot_dir)) dev.off()
    }
    if(!is.null(plot_dir)) png(glue::glue("{plot_dir}/hist_cor_df1_vs_df2.png"))
    hist(cor_vec,main="correlation between df1 and df2")
    if(!is.null(plot_dir)) dev.off()
  } else ## below is just testing, 10K by 10K is too large
  if(ncol(df1)*ncol(df2)<2e6){
    mat1 = scale( as.matrix(df1 %>% select(-c("FID","IID")) ) )
    mat2 = scale(as.matrix(df2 %>% select(-c("FID","IID")) ) )
    cor_mat = t(mat1) %*% mat2 / nrow(mat1)
  }
  list(df1,df2,cor_vec)
}

tempo_list = suppressWarnings(fn_compare_expression_matrices(df1, df2) )

```


Calculated correlation all genes agains all genes. Thsi cannot be done for all the genes, so I'm using a subset of 100 or so.

```{r testing all vs all correlation}
suppressMessages(library(wesanderson))



tileplot <- function(mat,no_label=FALSE)
{
  mat = data.frame(mat)
  mat$Var1 = factor(rownames(mat), levels=rownames(mat)) ## preserve rowname order
  melted_mat <- gather(mat,key=Var2,value=value,-Var1)
  melted_mat$Var2 = factor(melted_mat$Var2, levels=colnames(mat)) ## preserve colname order
  rango = range(melted_mat$value)
  melted_mat$val2 = cut(melted_mat$value,breaks=c(-1,-0.3,0.3,1))
  pp <- ggplot(melted_mat,aes(x=Var1,y=Var2,fill=val2)) + geom_tile() ##+scale_fill_gradientn(colours = c("#C00000", "#FF0000", "#FF8080", "#FFC0C0", "#FFFFFF", "#C0C0FF", "#8080FF", "#0000FF", "#0000C0"), limit = c(-1,1)) 
  if(no_label)  
    pp <- pp + 
    theme(axis.title.x=element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank(),
          axis.title.y=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank()) +
    scale_fill_manual(values = wes_palette("GrandBudapest1", n = 3))
          #scale_fill_brewer(palette = "Dark2")
          # scale_fill_gradientn( colours=rainbow(5),limit=c(-1,1) )
##    scale_fill_gradient( limits = c(-.5,.5),low =  "blue", high = "red", na.value="black") 
  print(pp)
}


  df1 = fn_remove_ver_from_names(df1)
  df2 = fn_remove_ver_from_names(df2)

t_genelist = sample(intersect(colnames(df1[,-c(1:2)]),colnames(df2[,-c(1:2)])),100)
    mat1 = scale( as.matrix(df1[,t_genelist]  ) )
    mat2 = scale(as.matrix(df2[,t_genelist]  ) )
cor_mat = t(mat1) %*% mat2 / nrow(mat1)
tileplot(cor_mat,no_label=TRUE)

```

